### 2018-03-14 submission/manuscript.Rmd
update intro paragraph 3 and 4

It is important to differentiate between bias, reproducibility, and standardization since often times these three can be confused and used interchangeably with each other. Bias can change the observed results in a way that is reproducible and standardized. For example, DNA extraction kits used for 16S rRNA gene sequencing biases your findings versus another group not using the same brand kit but both groups have reproducible results. Therefore, standardization of 16S rRNA gene sequencing methods to increase reproducibility can still be problematic due to bias. Determining how different aspects of a 16S rRNA gene sequencing workflow could bias the observed results is critical for the interpretation of specific studies within the context of the overall field.

A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps. Generally, not using a preservation media and leaving samples at room temperature has been shown to cause overgrowth of low abundance members of the fecal bacterial community [ref]. Similarily, this overgrowth could still occur if the preservation media does not adequately inhibit growth [ref]. Reports have also shown that changes in specific community members might occur due to differing susceptibility to freeze thaw cycles amongst microbes [ref]. Additionally, reagent contamination has been shown to add community members and the contribution of these contaminant members grows larger with lower biomass samples [@contamination_Salter2014]. Recent studies have shown that the majority of these biases due to either preservation or extraction tend to be smaller than the overall biological signal being measured [refs]. However, the contribution of PCR bias to this overall workflow is not well characterized since these studies use the same PCR approach while varying preservation or extraction method used.

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that primers used, cycle number, selective amplification, amplification error, non-specific amplification, chimeras, and polymerase dependent differences can all change the observed results [refs]. Many of these sources of biases are made worse as cycle number increases. The selective amplification of the dominant sequence can increase the difference between the most and next abundant sequence, over-emphasizing the actual difference [ref]. Both amplification error and non-specific amplification (e.g. incorrect amplicon size products) also increases as cycle number increases [refs]. Further, chimeras can form from an aborted extension followed by subsequent priming error and secondary extension [@Haas2011]. Even if it has not been explicitly shown to increase with cycle number it is easy to postulate that this does happen. There are also intrinsic properties to primers and DNA polymerases chosen that can introduce bias. With respect to primers, there are different binding affinities for different bacteria that are variable region dependent [refs]. Additionally, there are multiple families of DNA polymerases that have their own error rate and proof reading capcity [@polymerase_Ishino_2014]. Interestingly, the influence these different DNA polymerases can have on the observed results have not been well studied like the previously mentioned sources of PCR-based bias.

A recent study found clear differences between normal and high fidelity (HiFi) DNA polymerase and that you could reduce error and chimera generation by optimizing the PCR protocol [@taq_Gohl_2016]. This study also found that, regardless of DNA polymerase, the number of Operational Taxonomic Units (OTUs) or taxa generated were not easily reduced using the authors chosen bioinformatic pipeline  [@taq_Gohl_2016]. It is natural to extend this line of inquiry and ask if biases in the number of OTUs and chimeras are also dependent on HiFi DNA polymerases. There is some reason to think that this may be the case since many of these HiFi DNA polymerase come from different families (e.g. *Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014]. In this study we critically examine if any of five different HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

>>>>

This study will specifically address how cycle number and high fidelity (HiFi) DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing. It is important to differentiate between bias, reproducibility, and standardization since often times these three can be confused and used interchangeably with each other. Bias can change the observed results in a way that is reproducible and standardized. For example, if one group uses one brand of DNA extraction kit for their 16S rRNA gene sequencing, their results may be biased versus another group not using the same brand kit but within their group they can still have reproducible results. Therefore, standardization of 16S rRNA gene sequencing methods to increase reproducibility can still be problematic due to bias. Determining how different aspects of a 16S rRNA gene sequencing workflow could bias the observed results is critical for the interpretation of specific studies within the context of the overall field.

A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps. Generally, not using a preservation media and leaving samples at room temperature has been shown to cause overgrowth of low abundance members of the fecal bacterial community [ref]. Similarily, this overgrowth could still occur if the preservation media does not adequately inhibit growth [ref]. Reports have also shown that changes in specific community members might occur due to differing susceptibility to freeze thaw cycles amongst microbes [ref]. Additionally, reagent contamination has been shown to add community members and the contribution of these contaminant members grows larger with lower biomass samples [@contamination_Salter2014]. Recent studies have shown that the majority of these biases due to either preservation or extraction tend to be smaller than the overall biological signal being measured [refs]. However, the contribution of PCR bias to this overall workflow is not well characterized since these studies use the same PCR approach while varying preservation or extraction method.

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that primers used, cycle number, selective amplification, amplification error, non-specific amplification, chimeras, and polymerase dependent differences can all change the observed results [refs]. Many of these sources of biases are made worse as cycle number increases [ref]. For example, the selective amplification of the dominant sequence can increase the difference between the most and next abundant sequence, over-emphasizing the actual difference [ref]. Both amplification error and non-specific amplification (e.g. incorrect amplicon size products) can also increase as cycle number increases [refs]. Additionally, chimeras can form from an aborted extension followed by subsequent priming error and secondary extension and even if it has not been explicitly shown to increase with cycle number it is easy to postulate that this does happen [@Haas2011]. There are also intrinsic properties to primers and DNA polymerases chosen that can introduce bias. With respect to primers, there are different binding affinities for different bacteria that are dependent on variable region [refs]. Additionally, there are multiple families of DNA polymerases that have their own error rate and proof reading capcity [@polymerase_Ishino_2014]. Interestingly, the influence that these different DNA polymerases can have on the observed results have not been well studied like some of the other previously mentioned sources of PCR-based bias.

A recent study found clear differences between normal and high fidelity (HiFi) DNA polymerase and that optimization of the PCR protocol could reduce error and chimera generation [@taq_Gohl_2016]. This study also found that, regardless of DNA polymerase, the number of Operational Taxonomic Units (OTUs) or taxa generated were not easily reduced using the authors chosen bioinformatic pipeline [@taq_Gohl_2016]. It is natural to extend this line of inquiry and ask if biases in the number of OTUs and chimeras are also dependent on HiFi DNA polymerases. There is some reason to think that this may be the case since many of these HiFi DNA polymerase come from different families (e.g. *Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014]. In this study we critically examine if any of five different HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

### 2018-03-14 submission/manuscript.tex
update intro paragraph 3 and 4

It is important to differentiate between bias, reproducibility, and
standardization since often times these three can be confused and used
interchangeably with each other. Bias can change the observed results in
a way that is reproducible and standardized. For example, DNA extraction
kits used for 16S rRNA gene sequencing biases your findings versus
another group not using the same brand kit but both groups have
reproducible results. Therefore, standardization of 16S rRNA gene
sequencing methods to increase reproducibility can still be problematic
due to bias. Determining how different aspects of a 16S rRNA gene
sequencing workflow could bias the observed results is critical for the
interpretation of specific studies within the context of the overall
field.

A typical 16S rRNA gene sequencing workflow can be divided into
preservation, extraction, PCR, and sequencing steps. Generally, not
using a preservation media and leaving samples at room temperature has
been shown to cause overgrowth of low abundance members of the fecal
bacterial community {[}ref{]}. Similarily, this overgrowth could still
occur if the preservation media does not adequately inhibit growth
{[}ref{]}. Reports have also shown that changes in specific community
members might occur due to differing susceptibility to freeze thaw
cycles amongst microbes {[}ref{]}. Additionally, reagent contamination
has been shown to add community members and the contribution of these
contaminant members grows larger with lower biomass samples (Salter et
al., 2014). Recent studies have shown that the majority of these biases
due to either preservation or extraction tend to be smaller than the
overall biological signal being measured {[}refs{]}. However, the
contribution of PCR bias to this overall workflow is not well
characterized since these studies use the same PCR approach while
varying preservation or extraction method used.

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is
important because a large body of literature shows that primers used,
cycle number, selective amplification, amplification error, non-specific
amplification, chimeras, and polymerase dependent differences can all
change the observed results {[}refs{]}. Many of these sources of biases
are made worse as cycle number increases. The selective amplification of
the dominant sequence can increase the difference between the most and
next abundant sequence, over-emphasizing the actual difference
{[}ref{]}. Both amplification error and non-specific amplification
(e.g.~incorrect amplicon size products) also increases as cycle number
increases {[}refs{]}. Further, chimeras can form from an aborted
extension followed by subsequent priming error and secondary extension
(Haas et al., 2011). Even if it has not been explicitly shown to
increase with cycle number it is easy to postulate that this does
happen. There are also intrinsic properties to primers and DNA
polymerases chosen that can introduce bias. With respect to primers,
there are different binding affinities for different bacteria that are
variable region dependent {[}refs{]}. Additionally, there are multiple
families of DNA polymerases that have their own error rate and proof
reading capcity (Ishino \& Ishino, 2014). Interestingly, the influence
these different DNA polymerases can have on the observed results have
not been well studied like the previously mentioned sources of PCR-based
bias.

A recent study found clear differences between normal and high fidelity
(HiFi) DNA polymerase and that you could reduce error and chimera
generation by optimizing the PCR protocol (Gohl et al., 2016). This
study also found that, regardless of DNA polymerase, the number of
Operational Taxonomic Units (OTUs) or taxa generated were not easily
reduced using the authors chosen bioinformatic pipeline (Gohl et al.,
2016). It is natural to extend this line of inquiry and ask if biases in
the number of OTUs and chimeras are also dependent on HiFi DNA
polymerases. There is some reason to think that this may be the case
since many of these HiFi DNA polymerase come from different families
(e.g. \emph{Taq} belongs to the family A polymerases) and may
intrinsically have different error rates that cannot be completely
removed with modifications (Ishino \& Ishino, 2014). In this study we
critically examine if any of five different HiFi DNA polymerases
introduce significant biases into 16S rRNA gene surveys, if this is a
cycle dependent phenomenon, and whether they can be removed using a
standard bioinformatic pipeline.

>>>>

This study will specifically address how cycle number and high fidelity
(HiFi) DNA polymerases can bias observed bacterial community results
derived from 16S rRNA gene sequencing. It is important to differentiate
between bias, reproducibility, and standardization since often times
these three can be confused and used interchangeably with each other.
Bias can change the observed results in a way that is reproducible and
standardized. For example, if one group uses one brand of DNA extraction
kit for their 16S rRNA gene sequencing, their results may be biased
versus another group not using the same brand kit but within their group
they can still have reproducible results. Therefore, standardization of
16S rRNA gene sequencing methods to increase reproducibility can still
be problematic due to bias. Determining how different aspects of a 16S
rRNA gene sequencing workflow could bias the observed results is
critical for the interpretation of specific studies within the context
of the overall field.

A typical 16S rRNA gene sequencing workflow can be divided into
preservation, extraction, PCR, and sequencing steps. Generally, not
using a preservation media and leaving samples at room temperature has
been shown to cause overgrowth of low abundance members of the fecal
bacterial community {[}ref{]}. Similarily, this overgrowth could still
occur if the preservation media does not adequately inhibit growth
{[}ref{]}. Reports have also shown that changes in specific community
members might occur due to differing susceptibility to freeze thaw
cycles amongst microbes {[}ref{]}. Additionally, reagent contamination
has been shown to add community members and the contribution of these
contaminant members grows larger with lower biomass samples (Salter et
al., 2014). Recent studies have shown that the majority of these biases
due to either preservation or extraction tend to be smaller than the
overall biological signal being measured {[}refs{]}. However, the
contribution of PCR bias to this overall workflow is not well
characterized since these studies use the same PCR approach while
varying preservation or extraction method.

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is
important because a large body of literature shows that primers used,
cycle number, selective amplification, amplification error, non-specific
amplification, chimeras, and polymerase dependent differences can all
change the observed results {[}refs{]}. Many of these sources of biases
are made worse as cycle number increases {[}ref{]}. For example, the
selective amplification of the dominant sequence can increase the
difference between the most and next abundant sequence, over-emphasizing
the actual difference {[}ref{]}. Both amplification error and
non-specific amplification (e.g.~incorrect amplicon size products) can
also increase as cycle number increases {[}refs{]}. Additionally,
chimeras can form from an aborted extension followed by subsequent
priming error and secondary extension and even if it has not been
explicitly shown to increase with cycle number it is easy to postulate
that this does happen (Haas et al., 2011). There are also intrinsic
properties to primers and DNA polymerases chosen that can introduce
bias. With respect to primers, there are different binding affinities
for different bacteria that are dependent on variable region {[}refs{]}.
Additionally, there are multiple families of DNA polymerases that have
their own error rate and proof reading capcity (Ishino \& Ishino, 2014).
Interestingly, the influence that these different DNA polymerases can
have on the observed results have not been well studied like some of the
other previously mentioned sources of PCR-based bias.

A recent study found clear differences between normal and high fidelity
(HiFi) DNA polymerase and that optimization of the PCR protocol could
reduce error and chimera generation (Gohl et al., 2016). This study also
found that, regardless of DNA polymerase, the number of Operational
Taxonomic Units (OTUs) or taxa generated were not easily reduced using
the authors chosen bioinformatic pipeline (Gohl et al., 2016). It is
natural to extend this line of inquiry and ask if biases in the number
of OTUs and chimeras are also dependent on HiFi DNA polymerases. There
is some reason to think that this may be the case since many of these
HiFi DNA polymerase come from different families (e.g. \emph{Taq}
belongs to the family A polymerases) and may intrinsically have
different error rates that cannot be completely removed with
modifications (Ishino \& Ishino, 2014). In this study we critically
examine if any of five different HiFi DNA polymerases introduce
significant biases into 16S rRNA gene surveys, if this is a cycle
dependent phenomenon, and whether they can be removed using a standard
bioinformatic pipeline.
